fit_crcvm

### 2025-01-07 R/application/CRCVM/fit_crcvm.Rmd
R code updates

---
title: "fit_crcvm"
author: "Alexandre Delporte"
date: "2024-06-07"
output: html_document
---

@@

## Set up

```{r packages}
set.seed(42)

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
```

```{r get data}

# Set the path to the directory containing the data
par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")

#data before exposure with first 24h removed
dataBE1=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE2=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

@@

We choose thresholds on the distance to the shore to define the ExpShore covariate.

```{r preprocess data}

D_low=0.07
D_up=3
dataBE1[dataBE1$DistanceShore> D_up,"ExpShore"]=0
dataBE1[dataBE1$DistanceShore< D_low,"ExpShore"]=1/D_low

dataBE2[dataBE2$DistanceShore> D_up,"ExpShore"]=0
dataBE2[dataBE2$DistanceShore< D_low,"ExpShore"]=1/D_low

dataAE[dataAE$DistanceShore> D_up,"ExpShore"]=0
dataAE[dataAE$DistanceShore< D_low,"ExpShore"]=1/D_low

N=length(unique(dataBE2$ID))

n_pre=length(dataBE2$time) #number of observations before exposure
n_post=length(dataAE$time) #number of observations after exposure
```

## Baseline models

We can estimate the measurement error standard deviation directly from the data. Based on accuracy studies about fastloc GPS, it should be a few tens of metres. The estimate we get from the data before exposure is 35m (std error 1m), which seems consistent with previous studies. However, when estimating this measurement error from the post exposure data, we get a higher value of 48m, which is weird : the same GPS is device is used during all the experiment, so measurement error before or after exposure should be about the same. If we fix the measurement error to 35m when estimating the response, then we can't estimate because TMB gets a non finite value of the negative log-likelihood at initial parameters. This cannot be solved by changing the other initial parameters in the optimisation. Thus we decided to fix the measurement error to different values in the baseline and the response and compare the results we get. Higher measurement error gives higher estimates for tau, whereas the estimate of nu is not affected by the changes in the measurement error.

### Baselines with constant parameters

```{r fit baseline0}

#initial parameters
par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1,omega=~1)

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_100m$fit()

@@

```{r results baseline0}
estimates_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Est")
std_bas0_35m=as.list(baseline0_35m$tmb_rep(),what="Std")

estimates_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Est")
std_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Std")

estimates_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Est")
std_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Std")

estimates_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Est")
std_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Std")

@@

### Baseline with AngleNormal covariate in omega

Fixing a high standard deviation (>50m) for the measurement error tends to increase the estimate of the parameter tau. Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.

```{r fit baseline1}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1/0.2^2,1/0.2^2,0.01)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_35m$update_lambda(init_lambda)

baseline1_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_50m$update_lambda(init_lambda)

baseline1_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_75m$update_lambda(init_lambda)

baseline1_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_100m$update_lambda(init_lambda)

baseline1_100m$fit()

@@

```{r results baseline1}
estimates_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Est")
std_bas1_35m=as.list(baseline1_35m$tmb_rep(),what="Std")

estimates_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Est")
std_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Std")

estimates_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Est")
std_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Std")

estimates_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Est")
std_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Std")

plots_bas1_35m=baseline1_35m$get_all_plots(baseline=NULL,model_name="baseline1_35m",show_CI="pointwise",save=TRUE)
plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)

### Baseline with AngleNormal and DistanceShore in omega

```{r fit baseline2}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_35m$update_lambda(init_lambda)
baseline2_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_50m$update_lambda(init_lambda)
baseline2_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_75m$update_lambda(init_lambda)
baseline2_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_100m$update_lambda(init_lambda)
baseline2_100m$fit()

```{r get results baseline2}

estimates_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Est")
std_bas2_35m=as.list(baseline2_35m$tmb_rep(),what="Std")

estimates_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Est")
std_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Std")

estimates_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Est")
std_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Std")

estimates_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Est")
std_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("DistanceShore"=0.07)
xmax=list("DistanceShore"=5)

plots_bas2_35m=baseline2_35m$get_all_plots(baseline=NULL,model_name="baseline2_35m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
```

```{r estimates on parameter scale baseline2}

post_coeff_bas2_35m=baseline2_35m$post_coeff(n_post=1000)
post_par_bas2_35m=list("tau0"=exp(post_coeff_bas2_35m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_35m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_35m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_35m$log_lambda[,2])))
mean_par_bas2_35m=lapply(post_par_bas2_35m,mean)
sd_par_bas2_35m=lapply(post_par_bas2_35m,sd)

post_coeff_bas2_50m=baseline2_50m$post_coeff(n_post=1000)
post_par_bas2_50m=list("tau0"=exp(post_coeff_bas2_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,2])))
mean_par_bas2_50m=lapply(post_par_bas2_50m,mean)
sd_par_bas2_50m=lapply(post_par_bas2_50m,sd)

post_coeff_bas2_75m=baseline2_75m$post_coeff(n_post=1000)
post_par_bas2_75m=list("tau0"=exp(post_coeff_bas2_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,2])))
mean_par_bas2_75m=lapply(post_par_bas2_75m,mean)
sd_par_bas2_75m=lapply(post_par_bas2_75m,sd)

#get 95% confidence intervals
post_coeff_bas2_100m=baseline2_100m$post_coeff(n_post=1000)
post_par_bas2_100m=list("tau0"=exp(post_coeff_bas2_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,2])))
mean_par_bas2_100m=lapply(post_par_bas2_100m,mean)
sd_par_bas2_100m=lapply(post_par_bas2_100m,sd)
```

```{r smooth omega baseline2}

#get angular velocity for different distances to shore

plot1=plots_bas2_50m$fe_omega_AngleNormal_q1_DistanceShore
plot2=plots_bas2_50m$fe_omega_AngleNormal_q2_DistanceShore
plot4=plots_bas2_50m$fe_omega_AngleNormal_q4_DistanceShore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Close"
omega2$Category <- "Medium"
omega4$Category <- "Far"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Close"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Far"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
  theme_minimal()

ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline2_50m",width = 10, height = 6, units = "in")

### Baseline with AngleNormal and ExpShore in omega

```{r fit baseline3}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1,1,0.01,1,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~ti(AngleNormal,k=5,bs="cs")+ti(ExpShore,k=5,bs="cs")+ti(AngleNormal,ExpShore,k=c(5,5),bs="cs"))

# Fit baseline with 35m measurement error
sigma_obs=0.035
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_35m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_35m$update_lambda(init_lambda)
baseline3_35m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_50m$update_lambda(init_lambda)
baseline3_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_75m$update_lambda(init_lambda)
baseline3_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline3_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline3_100m$update_lambda(init_lambda)
baseline3_100m$fit()
```
```{r get results baseline3}

estimates_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Est")
std_bas3_35m=as.list(baseline3_35m$tmb_rep(),what="Std")

estimates_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Est")
std_bas3_50m=as.list(baseline3_50m$tmb_rep(),what="Std")

estimates_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Est")
std_bas3_75m=as.list(baseline3_75m$tmb_rep(),what="Std")

estimates_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Est")
std_bas3_100m=as.list(baseline3_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("ExpShore"=1/D_up)
xmax=list("ExpShore"=1/D_low)
link=list("ExpShore"=(\(x) 1/x))
xlabel=list("ExpShore"="Distance to shore")

plots_bas3_35m=baseline3_35m$get_all_plots(baseline=NULL,model_name="baseline3_35m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_50m=baseline3_50m$get_all_plots(baseline=NULL,model_name="baseline3_50m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_75m=baseline3_75m$get_all_plots(baseline=NULL,model_name="baseline3_75m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

plots_bas3_100m=baseline3_100m$get_all_plots(baseline=NULL,model_name="baseline3_100m",
                            xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise",save=TRUE)

```{r estimates on parameter scale baseline3}

post_coeff_bas3_35m=baseline3_35m$post_coeff(n_post=1000)
post_par_bas3_35m=list("tau0"=exp(post_coeff_bas3_35m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_35m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_35m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_35m$log_lambda[,2])))
mean_par_bas3_35m=lapply(post_par_bas3_35m,mean)
sd_par_bas3_35m=lapply(post_par_bas3_35m,sd)

post_coeff_bas3_50m=baseline3_50m$post_coeff(n_post=1000)
post_par_bas3_50m=list("tau0"=exp(post_coeff_bas3_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_50m$log_lambda[,2])))
mean_par_bas3_50m=lapply(post_par_bas3_50m,mean)
sd_par_bas3_50m=lapply(post_par_bas3_50m,sd)

post_coeff_bas3_75m=baseline3_75m$post_coeff(n_post=1000)
post_par_bas3_75m=list("tau0"=exp(post_coeff_bas3_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_75m$log_lambda[,2])))
mean_par_bas3_75m=lapply(post_par_bas3_75m,mean)
sd_par_bas3_75m=lapply(post_par_bas3_75m,sd)

post_coeff_bas3_100m=baseline3_100m$post_coeff(n_post=1000)
post_par_bas3_100m=list("tau0"=exp(post_coeff_bas3_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas3_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas3_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas3_100m$log_lambda[,2])))
mean_par_bas3_100m=lapply(post_par_bas3_100m,mean)
sd_par_bas3_100m=lapply(post_par_bas3_100m,sd)

```{r smooth omega baseline3}

plot1=plots_bas3_50m$fe_omega_AngleNormal_q1_ExpShore
plot2=plots_bas3_50m$fe_omega_AngleNormal_q2_ExpShore
plot4=plots_bas3_50m$fe_omega_AngleNormal_q4_ExpShore

omega1 <- ggplot_build(plot1)$data[[1]]
omega2 <- ggplot_build(plot2)$data[[1]]
omega4 <- ggplot_build(plot4)$data[[1]]

omega1$Category <- "Far"
omega2$Category <- "Medium"
omega4$Category <- "Close"

# Combine the data frames
combined_omega <- rbind(omega1, omega2, omega4)

ci_omega1 <- ggplot_build(plot1)$data[[2]]
ci_omega2 <- ggplot_build(plot2)$data[[2]]
ci_omega4 <- ggplot_build(plot4)$data[[2]]

ci_omega1$Category <- "Far"
ci_omega2$Category <- "Medium"
ci_omega4$Category <- "Close"

# Combine the data frames
combined_ci_omega <- rbind(ci_omega1, ci_omega2, ci_omega4)

combined_plot <- ggplot() +
  geom_line(data=combined_omega,aes(x=x,y = y,color = Category, linetype = Category)) +
  geom_ribbon(data=combined_ci_omega,aes(x=x,ymin=ymin,ymax=ymax,fill=Category),alpha=0.1)+
  labs(color = "Category", linetype = "Category",x="Theta",y="Omega") +
  geom_vline(xintercept = c(-pi/2, pi/2), linetype = "dashed", color = "grey")+
  annotate("text", x = -pi/2, y = Inf, label = expression(-pi/2), vjust = 40, color = "black") +
  annotate("text", x = pi/2, y = Inf, label = expression(pi/2), vjust = 40, color = "black")+
  theme_minimal()

ggsave("omega_DistanceShore_levels.png",plot=combined_plot,path="baseline3_50m",width = 10, height = 6, units = "in")

## Response models with splines of ExpShip

### Response models based on baseline1

```{r fit response1}

par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 nu=~s(ExpShip,k=3,bs="cs")+s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit response with 50m measurement error
sigma_obs=0.04
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_35m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_35m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_35m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_35m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_35m$lambda()["tau.s(ID)",1],1,baseline1_35m$lambda()["nu.s(ID)",1],
             baseline1_35m$lambda()["omega.s(AngleNormal)",1])
response1_sp_35m$update_coeff_re(new_coeff_re)
response1_sp_35m$update_coeff_fe(baseline1_35m$coeff_fe()[,1])

response1_sp_35m$update_lambda(new_lambda)

response1_sp_35m$fit()

# Fit response with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_50m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

new_coeff_re=c(rep(0,2),baseline1_50m$coeff_re()[paste("tau.s(ID).",1:N,sep=""),1],rep(0,2),
               baseline1_50m$coeff_re()[paste("nu.s(ID).",1:N,sep=""),1],
               baseline1_50m$coeff_re()[paste("omega.s(AngleNormal).",1:4,sep=""),1])
new_lambda=c(1,baseline1_50m$lambda()["tau.s(ID)",1],1,baseline1_50m$lambda()["nu.s(ID)",1],
             baseline1_50m$lambda()["omega.s(AngleNormal)",1])
response1_sp_50m$update_coeff_re(new_coeff_re)
response1_sp_50m$update_coeff_fe(baseline1_50m$coeff_fe()[,1])

response1_sp_50m$update_lambda(new_lambda)

response1_sp_50m$fit()

# Fit response with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response1_sp_75m<- SDE$new(formulas = formulas,data = dataAE,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(coeff_re=factor(c("tau.s(ExpShip).1","tau.s(ExpShip).2",rep(NA,N),
                                               "nu.s(ExpShip).1","nu.s(ExpShip).2",rep(NA,N),rep(NA,4))),coeff_fe=factor(rep(NA,5)),
                                                                                      log_lambda=factor(c("tau.s(ExpShip)",NA,
                                                                                                          "nu.s(ExpShip)",NA,NA))))

>>>>

---
title: "fit_crcvm"
author: "Alexandre Delporte"
date: "2024-12-12"
output: html_document
---
To circumvent this issue, we can fix a smooth parametric function for $\omega$, based on the observed distances to shore and the observed angular velocities.

# Set up

@@

```{r packages,message=FALSE}
set.seed(42)

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate) # animated trajectories
```

```{r get data,message=FALSE}

# Set the path to the directory containing the data
par_dir=here() #parent directory 
narwhal_data_path <- file.path(par_dir,"Data","preprocessed_data","narwhals")

#data before exposure with first 24h removed
dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

@@

N=length(unique(dataBE12$ID))

n_pre12=length(dataBE12$time) #number of observations before exposure
n_pre24=length(dataBE24$time) #number of observations before exposure
n_post=length(dataAE$time) #number of observations after exposure
```

plot_Dshore=ggplot(data=dataBE12)+geom_point(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.5)+
  geom_line(aes(x=time,y=DistanceShore,col=ID),size=1,alpha=0.2)+
  theme_minimal()

ggsave("observed_Dshore.pdf",plot=plot_Dshore)

quantile(dataBE12$DistanceShore,probs=0.05)

@@

```{r}
dataBE12$angular_velocity=c(NA,diff(dataBE12$AngleNormal)/diff(dataBE12$time))
plot_angular_velocity=ggplot(data=dataBE12)+geom_point(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.5)+
  geom_line(aes(x=time,y=angular_velocity,col=ID),size=1,alpha=0.2)+
  theme_minimal()
ggsave("observed_angular_velocity.pdf",plot=plot_angular_velocity)
max(dataBE12$angular_velocity,na.rm=TRUE)/(pi*(pi-pi/2)*(pi+pi/2)*exp(-1))

@@

renamed_dataBE12=dataBE12
colnames(renamed_dataBE12)[48]="theta"
renamed_dataBE12[renamed_dataBE12$DistanceShore==0,]=0.02

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre12),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~s(theta,k=6,bs="cs"),D0=~1)

# Fit baseline
baseline_crcvm_cubic<- SDE$new(formulas = formulas,data = renamed_dataBE12,type = "CRCVM_SSM",
                    response = c("x","y"),par0 = c(2,4,0,0.05),other_data=list("H"=H),
                    fixpar=c("D0"))

baseline_crcvm_cubic$update_lambda(c(exp(3),exp(3),10^-12))
baseline_crcvm_cubic$update_map(list("log_lambda"=factor(c(1,2,NA))))
baseline_crcvm_cubic$fit()

@@

```{r}
plots_bas_crcvm=baseline_crcvm_cubic$get_all_plots(baseline=NULL,show_CI="simultaneous")

plots_bas_crcvm
```
